---
title: Fleeing
author: 
  - admin
date: '2020-08-28'
slug: fatal-police-shooting-and-fleeing
categories:
  - R
tags:
  - plot
  - R Markdown
  - flee
  - time
subtitle: ''
summary: ''
authors: []
lastmod: '2020-08-28T13:25:40-07:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)

library(modelr)
library(infer)
library(knitr)
library(broom)
library(lubridate)
library(tidyverse)
library(multcompView)
library(gganimate)
library(gapminder)
library(DT)
library(widgetframe)
library(htmlwidgets)
library(htmltools)

shootings <- read.csv("https://raw.githubusercontent.com/washingtonpost/data-police-shootings/master/fatal-police-shootings-data.csv", header = T)

shootings <- shootings %>% 
  mutate(date = lubridate::ymd(date)) %>%   # Convert date to Lubridate
  mutate(age = round(age)) %>%              # Make all ages integers
  mutate(gender = as.factor(gender),        # Factorize
         race = as.factor(race),
         signs_of_mental_illness = as.factor(signs_of_mental_illness),
         threat_level = as.factor(threat_level),
         flee = as.factor(flee),
         body_camera = as.factor(body_camera) )
shootings <- shootings %>%
  # Replace abbreviated race identifier with race
  mutate(race = as.factor( ifelse(race=="W","White",
                  ifelse(race=="B","Black",
                  ifelse(race=="H", "Hispanic",
                  ifelse(race=="A", "Asian",
                  ifelse(race=="N", "Native",
                  ifelse(race=="O", "Other", "Not Recorded")))))))) %>%
  mutate(year = year(date)) %>%
  mutate(flee = as.factor( ifelse(flee=="Car","Car",
                  ifelse(flee=="Foot","Foot",
                  ifelse(flee=="Not fleeing", "Not Fleeing",
                  ifelse(flee=="Other", "Other", "Not Recorded"))))))
```



This analysis is going to explore some of the different circumstances behind fatal police shootings in where the victim was fleeing.

# Type of Fleeing 
Figure 1 below shows the proportion of those who do not flee, who do, and if they are fleeing, what method of fleeing that these individuals are using. 

```{r}
#proportion are killed while attempting to flee
shootings %>%
  mutate(count = 1) %>%
  aggregate(count ~ flee + year, data = ., sum) %>%
  group_by(year) %>%
  mutate(proportion = count / sum(count)) %>%
  ggplot() +
    geom_line(mapping = aes(x = year, y = proportion, color = flee)) +
    ggtitle("Proportion of Fatal Police Shooting by Flee Method") + 
    xlab("Year") + ylab("Proportion") + theme_bw()

```
Figure 1.
*The proportion of fatal police shootings by flee method over time.*

Figure 1 depicts the proportion of fleeing by car decreasing and every other method is seen increasing, in recent. The proportion of the incidents where fleeing method is not recorded is increasing every year, almost to the same degree that the "Not Fleeing" method is decreasing, this should be taken into account. 




```{r}
shootings %>%
  mutate(count = 1) %>%
  filter(armed == "unarmed") %>%
  aggregate(count ~ flee + year + armed, data = ., sum) %>%
  group_by(year) %>%
  mutate(proportion = count / sum(count)) %>%
  ggplot(mapping = aes(x=year, y = proportion, color = flee)) +
  geom_line() +
  ggtitle("Breakdown of Unarmed Victims") + xlab("Year") + ylab("Proportion") + theme_bw() 

```
Figure 2. *Accounts for the proportion of method of fleeing of unarmed victims of fatal police shootings*

```{r}
unarmed <- shootings %>%
  mutate(count = 1) %>%
  filter(armed == "unarmed") %>%
  aggregate(count ~ flee + year + armed, data = ., sum) %>%
  group_by(year) %>%
  mutate(proportion = count / sum(count)) %>%
  select(-armed, -count) %>%
  datatable()


htmlwidgets::saveWidget(unarmed, file = "unarmed.html", selfcontained = TRUE)

```
<iframe seamless src="../unarmed.html" width="100%" height="500"></iframe>

Table 1. 
*A table of the proportion of unarmed victims by fleeing method*


