---
title: Exploration of Mental Illness and Fatal Police Shootings
author: 
- admin
date: '2020-08-26'
slug: exploration-of-mental-illness-and-fatal-police-shootings
categories:
  - R
tags:
  - R Markdown
  - plot
subtitle: ''
summary: ''
authors: []
lastmod: '2020-08-26T22:40:23-07:00'
featured: false
draft: false
image:
  caption: ''
  focal_point: ''
  preview_only: false
projects: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)

library(modelr)
library(infer)
library(knitr)
library(broom)
library(lubridate)
library(tidyverse)
library(multcompView)
library(gganimate)
library(gapminder)
library(DT)
library(widgetframe)
library(htmlwidgets)
library(htmltools)

shootings <- read.csv("https://raw.githubusercontent.com/washingtonpost/data-police-shootings/master/fatal-police-shootings-data.csv", header = T)

```

# Mental Illness

There's a statistic that people mental illness are 15 times more likely to be shot in a police encounter than the lay person. We can think about people like Shukri Ali Said who's sister called 911 for aid when Said had a mental break. Police had found Said at an intersection with a knife and shot her 5 times, resulting in her death. We can think about Saheed Vassell. A man with a recorded mental illness, who also had a fatal police shooting. Vassell pointed a pipe at a police officer and was shot. We can think about the many people with mental and intellectual illnesses who had fatal police encounters and think about the possible different roots of the problem; the history of social stigma of treatment, to the processes of deinstitutionalization, to the defunding of existing community services like education, affordable housing, social and health services. 

This post is going to explore the effects of mental illness and fatal police encounters. 

# General Statistics

You can find the dataset [here](https://github.com/washingtonpost/data-police-shootings).

```{r include=FALSE}

shootings <- shootings %>% 
  mutate(date = lubridate::ymd(date)) %>%   # Convert date to Lubridate
  mutate(age = round(age)) %>%              # Make all ages integers
  mutate(gender = as.factor(gender),        # Factorize
         race = as.factor(race),
         signs_of_mental_illness = as.factor(signs_of_mental_illness),
         threat_level = as.factor(threat_level),
         flee = as.factor(flee),
         body_camera = as.factor(body_camera) )
shootings <- shootings %>%
  # Replace abbreviated race identifier with race
  mutate(race = as.factor( ifelse(race=="W","White",
                  ifelse(race=="B","Black",
                  ifelse(race=="H", "Hispanic",
                  ifelse(race=="A", "Asian",
                  ifelse(race=="N", "Native",
                  ifelse(race=="O", "Other", "Not Recorded")))))))) %>%
  mutate(year = year(date))


```




```{r echo=FALSE, message=FALSE, warning=FALSE}
mental.illness <- shootings %>%
  mutate(count = 1) %>%
  aggregate(count ~  race + signs_of_mental_illness + year, data = ., sum) %>%
  group_by(race, signs_of_mental_illness) %>%
  summarise(mean = mean(count),
            sd = sd(count),
            min = min(count),
            max = max(count)) 
d1 <- mental.illness %>%
  datatable()


  

```




```{r include=FALSE}
htmlwidgets::saveWidget(d1, file = "d1.html", selfcontained = TRUE)
```



<iframe seamless src="../d1.html" width="100%" height="500"></iframe>

Table 1
*General descriptive statistics of fatal police shootings overall.*


```{r echo=FALSE, message=FALSE, warning=FALSE}
ag.year <- shootings %>%
  mutate(count = 1) %>%
  aggregate(count ~  race + signs_of_mental_illness + year, data = ., sum) %>%
  group_by(race, signs_of_mental_illness, year) %>%
  summarise(mean = mean(count))
  
d2 <- ag.year %>%
  datatable()


ag.year %>%
  ggplot() +
  geom_point(mapping = aes(x = race, y = mean), color = "gray") +
  geom_boxplot(mapping = aes(x = race, y = mean, fill = signs_of_mental_illness)) +
  theme_bw() + ggtitle("Average Shootings by Mental Illness") + xlab("Race") + ylab("Average")
```
Figure 1
*Boxplot of the mean fatal police encounters by race over the years.*
```{r include=FALSE}
htmlwidgets::saveWidget(d2, file = "d2.html", selfcontained = TRUE)
```

<iframe seamless src="../d2.html" width="100%" height="500"></iframe>

Table 2
*Mean of police shooting by race aggregated by year.*


```{r eval=FALSE, include=FALSE}
shootings %>%
  mutate(count=1) %>%
  aggregate(count ~ signs_of_mental_illness + race, data = ., sum) %>%
  group_by(race) %>%
  mutate(prop = count / sum(count))
```

In the two tables shown above, we get a good picture fatal police shootings by signs of mental illness, from the years 2015 to current day. White people make up more than 70% of the US population, so it makes sense that white people will make up the most of police shootings in just raw numbers. Look at Figure 1 for the example. If we do some data crunching, we wil see something interesting. Since 2015, even though white people make up more than 70% of the population, 45% of fatal shootings involve white victims, which is disproportionate. If we factor in mental illness, overall, almost 30% of the White victims of fatal police shootings showed signs of mental illness. Of Black victims: 14.3%, of Asian: 24.7%, of Hispanic: 17.2%, of Native American: 17.9%. Overall, 28.9% of fatal police shooting victims are recorded as showing signs of mental illness. 


# De-escalation?

If more than 1 in 4 victims of fatal police shootings involve mental illness, were there attempts to subdue these victims besides shooting? The dataset I'm working with involve whether a taser was used to de-escalate along with shooting via the variable `manner_of_death`. There were only two categories, `shot`, and, `shot and Tasered`.


```{r echo=FALSE}
shootings %>%
  filter(signs_of_mental_illness == "True") %>%
  mutate(count = 1) %>%
  aggregate(count ~ manner_of_death + year, data = ., sum) %>%
  group_by(year) %>%
  mutate(prop = count / sum(count)) %>%
  ggplot() +
  geom_line(mapping = aes(x = year, y = prop, color = manner_of_death)) +
  theme_bw() + xlab("Year") + ylab("Proportion") + ggtitle("Victims with Mental Illness")
```
Figure 2.


```{r eval=FALSE, include=FALSE}
shootings %>%
  filter(signs_of_mental_illness == "True") %>%
  mutate(count = 1) %>%
  aggregate(count ~ manner_of_death + year, data = ., sum) %>%
  group_by(year) %>%
  mutate(prop = count / sum(count))
```
In Figure 2, we see that with the fatal police shootings involving victims that showed signs of mental illness, less than 10% in every year have on record that the police tried to deploy a taser as a means of subduing the victim. we actually see a dip in 2018 and is slowly starting to rise again. We are only through a little more than half of the year, but this dataset is updated real time, so this graph will always be current.


```{r echo=FALSE, message=FALSE, warning=FALSE}
shootings %>%
  filter(signs_of_mental_illness == "True", manner_of_death == "shot and Tasered") %>%
  mutate(count = 1) %>%
  aggregate(count ~ manner_of_death + year + race, data = ., sum) %>%
  group_by(year) %>%
  mutate(prop = count / sum(count)) %>%
  ggplot() +
  geom_smooth(mapping = aes(x = year, y = prop, color = race)) +
  theme_bw() + xlab("Year") + ylab("Proportion") + ggtitle("Victims with Mental Illness by Race")
```
Figure 3
*Taser deployed with victims showing signs of mental illness by race*


```{r eval=FALSE, include=FALSE}
shootings %>%
  filter(signs_of_mental_illness == "True", manner_of_death == "shot and Tasered") %>%
  mutate(count = 1) %>%
  aggregate(count ~ manner_of_death + race, data = ., sum) %>%
  #group_by(race) %>%
  mutate(prop = count / sum(count))
```



Figure 3 offers more insight. Overall, through all years, White victims that involve deployment of a taser: 47.8%, Native American: 1.11%, Hispanic: 17.8%, Black: 23.3%, Asian: 3.3%. We also see that the proportion of the use and deployment of a taser is more steadily increasing among black and white victims, those who make up the majority of victims of fatal police shootings. The percentage of hispanic victims look to be increasing to a small degree, but it may be attributed to the decrease in proportion of victims with race not being recorded.


# Conclusion

With almost 28.9% of all fatal police shootings involve the victims showing signs of mental illness. Of those instances, less than 10% every year are recorded that other means than just shooting a gun were used to subdue the victim. Figure 3 shows the ratio of those instances by race with white and black percentages increasing since 2018. 

It still begs the question. If 28.9% of victims of fatal police shootings show signs of mental illness, why do 10% of those cases have other means besides shooting as a way to subdue these individuals? I do think it is important to note that certain states have banned the use of tasers. However, could these instances of mental breaks or threatening individuals with mental illnesses been avoided all together? Should we have invested more in social services and health services? These are questions that can't be answered with this dataset alone. 



