---
title: Threat and Race
author: 
- admin
date: '2020-08-28'
slug: threat-and-race
categories:
  - R
tags:
  - R Markdown
  - plot
  - ANOVA
  - time
subtitle: ''
summary: ''
authors: []
lastmod: '2020-08-28T16:00:15-07:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)

library(modelr)
library(infer)
library(knitr)
library(broom)
library(lubridate)
library(tidyverse)
library(multcompView)
library(gganimate)
library(gapminder)
library(DT)
library(widgetframe)
library(htmlwidgets)
library(htmltools)

shootings <- read.csv("https://raw.githubusercontent.com/washingtonpost/data-police-shootings/master/fatal-police-shootings-data.csv", header = T)

shootings <- shootings %>% 
  mutate(date = lubridate::ymd(date)) %>%   # Convert date to Lubridate
  mutate(age = round(age)) %>%              # Make all ages integers
  mutate(gender = as.factor(gender),        # Factorize
         race = as.factor(race),
         signs_of_mental_illness = as.factor(signs_of_mental_illness),
         threat_level = as.factor(threat_level),
         flee = as.factor(flee),
         body_camera = as.factor(body_camera) )
shootings <- shootings %>%
  # Replace abbreviated race identifier with race
  mutate(race = as.factor( ifelse(race=="W","White",
                  ifelse(race=="B","Black",
                  ifelse(race=="H", "Hispanic",
                  ifelse(race=="A", "Asian",
                  ifelse(race=="N", "Native",
                  ifelse(race=="O", "Other", "NR")))))))) %>%
  mutate(year = year(date)) %>%
  mutate(flee = as.factor( ifelse(flee=="Car","Car",
                  ifelse(flee=="Foot","Foot",
                  ifelse(flee=="Not fleeing", "Not Fleeing",
                  ifelse(flee=="Other", "Other", "Not Recorded"))))))
```

# Threat and Race
There has always been a debate between the links of fatal police shootings and victims. There has even been insinuations that race has an effect on the threatening nature of these individuals. This topic often sparks outrage within the black and POC communities and within law enforcement. 

I'm going to try to explore this sensitive topic and try to make some inferential findings.

# The Data
The dataset can be found [here](https://github.com/washingtonpost/data-police-shootings).

We have three variables to look at: "attack", "other", and "undetermined". Attack constitutes as the individual was reported to have posed immediate threat to the officer. "Other" is every other level of threat and "undetermined" is for any circumstance that through the vetting process, threat level could not be determined. 
Keep in mind how the level of threat is determined, as explained in the home page.

Let's look at the visualization of the raw numbers/counts. Figure 1 below shows the level of reported threat by race. 

```{r}
shootings %>%
  mutate(count = 1) %>%
  ggplot() +
  geom_bar(mapping = aes(x = threat_level, fill = race), position = "dodge") +
  facet_wrap(~threat_level, scales = "free") + 
  ggtitle("Raw Amounts of Threat Level") +
  xlab("Threat Level") +
  ylab("Count")
```
Figure 1.
*Amount of Fatal Police Shootings by Race since 2015* 

All three graphs look identical. The only visual difference is the volume of the counts. I think in this case, looking at the proportions will be more inferenital.

```{r}
r.proportion <- shootings %>%
  mutate(count = 1) %>%
  mutate(year = year(date)) %>%
  aggregate(count ~ threat_level + race + year, data = ., sum) %>%
  filter(threat_level == "attack") %>%
  group_by(year) %>%
  mutate(race.proportion = count / sum(count))
r.proportion %>%
  ggplot() +
  geom_boxplot(mapping = aes(x = race, y = race.proportion), fill = "lavender") +
  geom_point(mapping = aes(x = race, y = race.proportion, color = year)) +
  ggtitle("Proportion of Race with 'attack' Threat Level") + xlab("Race") +
  ylab("Proportion") + theme_bw()


```
Figure 2. 
*Box plot and distribution of the proportions of victims who are recorded to have posed a threat to an officer's life by race from 2015 to current.*

```{r}
r.proportion %>%
  lm(formula = race.proportion ~ race * year, data = .) -> model
r.proportion
r.proportion %>%
  ggplot(aes(x= race, y = race.proportion, color = year)) +
    geom_point() +
    stat_smooth(method = "lm")
summary(model)
```

```{r}
ANOVA.threat = aov(model)

# Tukey test to study each pair of treatment :
TUKEY <- TukeyHSD(x=ANOVA.threat, 'race', conf.level=0.95, ordered = TRUE)
# Tuckey test representation :
plot(TUKEY , las=1 , col="brown")


```



```{r}
TUKEY
```



```{r}
# I need to group the treatments that are not different each other together.
label_df <- function(TUKEY, variable){
 
     # Extract labels and factor levels from Tukey post-hoc 
     levels <- TUKEY[[variable]][,4]
     Tukey.labels <- data.frame(multcompLetters(levels)['Letters'])
     
     #I need to put the labels in the same order as in the boxplot :
     Tukey.labels$treatment=rownames(Tukey.labels)
     Tukey.labels=Tukey.labels[order(Tukey.labels$treatment) , ]
     return(Tukey.labels)
     }
 
# Apply the function on my dataset
LABELS <- label_df(TUKEY , "race")
 
 
# A panel of colors to draw each group with the same color :
my_colors <- c( 
  rgb(143,199,74,maxColorValue = 255),
  rgb(242,104,34,maxColorValue = 255), 
  rgb(111,145,202,maxColorValue = 255),
  rgb(151,105,102,maxColorValue = 255)
  )
 
# Draw the basic boxplot
a <- boxplot(r.proportion$race.proportion ~ r.proportion$race, ylim=c(min(r.proportion$race.proportion) , 1.1*max(r.proportion$race.proportion)) , col=my_colors[as.numeric(LABELS[,1])] , ylab="value" , main="")


# I want to write the letter over each box. Over is how high I want to write it.
over <- 0.1*max( a$stats[nrow(a$stats),] )
 
#Add the labels
text( c(1:nlevels(r.proportion$race)) , a$stats[nrow(a$stats),]+over , LABELS[,1]  , col  = my_colors[as.numeric(LABELS[,1])] )
```




