---
title: Fleeing
author: 
  - admin
date: '2020-08-28'
slug: fatal-police-shooting-and-fleeing
categories:
  - R
tags:
  - plot
  - R Markdown
  - flee
  - time
subtitle: ''
summary: ''
authors: []
lastmod: '2020-08-28T13:25:40-07:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)

library(modelr)
library(infer)
library(knitr)
library(broom)
library(lubridate)
library(tidyverse)
library(multcompView)
library(gganimate)
library(gapminder)
library(DT)
library(widgetframe)
library(htmlwidgets)
library(htmltools)

shootings <- read.csv("https://raw.githubusercontent.com/washingtonpost/data-police-shootings/master/fatal-police-shootings-data.csv", header = T)

shootings <- shootings %>% 
  mutate(date = lubridate::ymd(date)) %>%   # Convert date to Lubridate
  mutate(age = round(age)) %>%              # Make all ages integers
  mutate(gender = as.factor(gender),        # Factorize
         race = as.factor(race),
         signs_of_mental_illness = as.factor(signs_of_mental_illness),
         threat_level = as.factor(threat_level),
         flee = as.factor(flee),
         body_camera = as.factor(body_camera) )
shootings <- shootings %>%
  # Replace abbreviated race identifier with race
  mutate(race = as.factor( ifelse(race=="W","White",
                  ifelse(race=="B","Black",
                  ifelse(race=="H", "Hispanic",
                  ifelse(race=="A", "Asian",
                  ifelse(race=="N", "Native",
                  ifelse(race=="O", "Other", "Not Recorded")))))))) %>%
  mutate(year = year(date)) %>%
  mutate(flee = as.factor( ifelse(flee=="Car","Car",
                  ifelse(flee=="Foot","Foot",
                  ifelse(flee=="Not fleeing", "Not Fleeing",
                  ifelse(flee=="Other", "Other", "Not Recorded"))))))
```



# Fleeing and Some Implications
It's never really surprising to hear about a police chase in where the suspect gets shot and then passes away. It's not surprising when we hear of people fleeing from the police when the situation escalates. What is surprising to hear is when a law enforcement officer is prosecuted for the death of a fleeing sispect. In Georgia, a police officer had a fatal encounter with [a man who was fleeing](https://apnews.com/ae503a6672a34aff873260456224652f/Manslaughter-charge:-Georgia-officer-shot-fleeing-black-man) was charged with manslaughter. A police officer was charged for the shooting and death of a [teenager who fleed after a traffic stop](https://apnews.com/f721e40873ef4b06ac2c5e6c0200d0f8/Officer-charged-in-death-of-black-teen-who-was-shot-in-back). 

This analysis is going to explore some of the different circumstances behind fatal police shootings in where the victim was fleeing.

# Type of Fleeing 
Figure 1 below shows the proportion of those who do not flee, who do, and if they are fleeing, what method of fleeing that these individuals are using. 

```{r}
#proportion are killed while attempting to flee
shootings %>%
  mutate(count = 1) %>%
  aggregate(count ~ flee + year, data = ., sum) %>%
  group_by(year) %>%
  mutate(proportion = count / sum(count)) %>%
  ggplot() +
    geom_line(mapping = aes(x = year, y = proportion, color = flee)) +
    ggtitle("Proportion of Fatal Police Shooting by Flee Method") + 
    xlab("Year") + ylab("Proportion") 

```
Figure 1
*The proportion of fatal police shootings by flee method over time.*



leaving this for right 





